Blame and Coercion: Together Again for the Second Time

### 2019-10-17 paper.tex
rewrite intro

Around 2006, many work on integrating dynamic typing and static typing emerge, 
notably gradual types \citep{siek2006gradual} and hybrid types 
\citep{flanagan2006hybrid}. 
Later \citet{wadler2009well} introduces Blame Calculus, an intermediate 
language for gradual types and hybrid types. 
Blame Calculus throws a blame label when a runtime type-check fails.
Labels provide useful debugging information, therefore are necessary for 
practical use of gradually typed languages.

Implementing gradually typed languages naively on top of blame calculus, 
however, suffices sever space leak.
For example, translation to Blame Calculus can wrap some expressions with code 
that performs runtime type-checking (casts), including expressions at 
tail positions. Thus, mutually tail-recursive functions in source language can 
consume unbounded space at runtime. There are more examples of space leaking in 
\citet{herman2010space}.

In 2007, \citet{herman2010space} proposes a solution to this problem by 
representing casts with coercion \citep{henglein1994dynamic}, 
which can be composed and normalized. Every coercion looks like a list of 
trees. For normal coercion, the length of the list is bounded by a small 
constant. And the depths of trees do not grow during composition and
normalization. These two facts implies that coercion is a space-efficient cast 
representation. This coercion, however, does not support blame 
tracking.

After that, many efforts have been made to combine blame tracking and 
space efficiency.

\citet{siek2009exploring} present another intermediate langauge for gradually 
typed language, cast calculus \castCalculus\ and achieve both blame-tracking 
and space-efficiency for this language by decorating \citet{herman2010space}'s 
coercions with labels. \castCalculus\ is simply typed lambda calculus extended 
with the dynamic type and cast expressions. The labels in \castCalculus{}, 
unlike their counterparts in the Blame Calculus, have no polarity. 
They also propose that there are four blame strategies for \castCalculus{}:
$ \{Lazy, Eager\} \times \{D, UD\} $. Eager strategies detect more potential 
type errors, but blame more programs. $ D $ and $ UD $ assign blame labels 
differently. They claim that their \lazyUD\ cast calculus is the same as the 
Blame Calculus, modulo label polarity, and prove that their their \lazyUD\ 
coercion calculus simulates their \lazyUD{} cast calculus. It is unknown, 
however, whether the simulation in the other direction also works.
\todo[inline]{Is it important to this paper whether there is a
simulation in the other direction? -JS
\\ --- \\
I guessed to say coercion is correct the other direction is 
necessary. -KC}

\citet{siek2010threesomes} proposes another approach to combine blame 
tracking and space efficiency. Their solution represents casts with threesomes, 
a novel cast representation. Threesome is proved space-efficient as well.
They prove that their calculus bisimulates the \lazyUD{} coercion 
calculus in \citet{siek2009exploring} and the Blame Calculus in 
\citet{wadler2009well}.

\citet{siek2012interpretations} present a revised coercion calculi.
They simplify the coercion calculi by only 
working with normal coercions (i.e. coercions in normal form). 
They conjecture that their revised coercion calculi bisimulate the 
corresponding cast calculi.

\citet{garcia2013calculating} develops
threesome for other blame strategies. After studying in depth the connection 
between coercion and threesome,
he claims that normal coercion with labels is easy to understand but hard to 
implement and hard to reason about, and that 
threesome with labels, however, is easy to implement and easy to reason about 
but hard to understand. 
His claim is mostly affirmed by the group of people who develop threesome 
(\citet{siek2015blame}).
His threesome calculi are equivalent to the coercion calculi of
\citet{siek2009exploring} because the former ones are derived from the latter 
ones.

\citet{siek2015blame} revisit the coercion-based approach. They revise
the \lazyUD{} coercion in \citet{siek2012interpretations} and prove
that their revised coercion calculus bisimulates the \lazyUD{} Blame
Calculus \citet{wadler2009well}.

Last year, \citet{kuhlenschmidt2018efficient} present Grift, a compiler for a 
gradually typed language of the same name. 
This implementation is based on the \lazyD{} coercion in 
\citet{siek2012interpretations}.
Their result suggests that normal coercion is easy enough to implement in a 
compiler.

So far, it seems that normal coercion is the best way to combine 
space-efficiency and blame-tracking: normal coercion is claimed easy to 
understand \citep{garcia2013calculating}\citep{siek2015blame} and is shown 
easy enough to implement in a compiler \citep{kuhlenschmidt2018efficient}.
Normal coercion is still not satisfactory, however, in at least two aspects.
Firstly, as we mentioned above, it is hard to reason about normal coercion. The 
major difficulty is from its non-structurally recursive composition.
For instance, the developers of Grift report it is tricky to convince
their proof assistants that the composition terminates.

Secondly, the definition of normal coercion lies unnecessarily on top 
of coercion: all coercions are normal when they are initially constructed, and 
\citet{siek2012interpretations} have shown an function that 
composes and produces normal coercions.
%
\todo[inline]{Many researchers would view this as a positive thing,
  not a negative. For example, the notion of normal forms comes up
  over and over again in mathematics and logic. -JS
\\ --- \\
I guess having a notion of normal forms is more convenient when a 
notion of normalization is convenient. But we don't even need a notion of 
normalization for cast representations. -KC
\\ --- \\
FWIW, I agree with you. But the point is that the argument is not
likely to be convincing to the reader, so we're probably better off
not wasting space on it. -JS
}

Perhaps unsurprisingly, threesome has a structurally recursive compose and a 
self-standing definition. Together with \citet{garcia2013calculating}'s 
work, they suggest that there might be a cast representation whose definition 
is self-standing, and whose composition is structurally recursive. 
supercoercion introduced in \citet{garcia2013calculating} could be a promising 
candidate. It is believed even more complicated than threesome 
\citep{siek2015blame}. In fact, it uses 10 constructors to deal with an 
elementary type system with only base types and 
function types. And four constructors are directly related to function types. 
Thus, supercoercion might not scale very well to more sophisticated type 
systems.

We present hypercoercion, a cast representation whose composition is 
structurally recursive, and whose definition is self-standing. What's more, it 
should be at 
least as easy to implement and understand as normal coercions, because there is 
a clear connection between them, as we will show in Section 
\ref{sec:hypercoercion}.
Our hypercoercion considers sum types and product types, which are not 
considered in all proofs above. Adding each of them requires us to add only one 
new constructor to a component of hypercoercion. This suggests that 
hypercoercion might scale better than supercoercion.

We prove formally that the \lazyD{} (resp. \lazyUD{}) hypercoercion calculus 
bisimulates the \lazyD{} (resp. \lazyUD{}) cast calculus. This is 
the first bisimulation proof for the \lazyD{} cast calculus as far as we know.
We care more about lazy strategies because recently, \citet{new2019gradual} 
show that \eager{} strategies are incompatible with $\eta$-equivalence of 
functions, which suggest that these strategies are not very ideal. 
\todo[inline]{Explain why we care about \lazyD{} and \lazyUD{}. -JS
  \\ --- \\ Explained, not sure if it suffice. -KC
  \\ --- \\
  That's good that you explained why lazy instead of eager.
  (Also, you should add that we are interested in lazy because
  that's what Grift does.)\\
  But you haven't explained why we are interested in both D and UD. -JS
}

The structure of this paper is as follows. 
Section \ref{sec:blame-calculus} reviews the state-of-art of $Lazy$ cast 
calculi. 
Section \ref{sec:coercion-calculus} reviews the state-of-art of $Lazy$ Coercion 
Calculi. 
In section \ref{sec:hypercoercion} we present hypercoercion.
Section \ref{sec:conclude} concludes.
\todo[inline]{Lots of missing sections. -JS}

\section{Background} \label{sec:blame-calculus}

@@

\section{hypercoercion} \label{sec:hypercoercion}

@@

\section{Space-efficient CEK Abstract Machine}
\label{sec:CEKS}

@@

\section{Correctness Proof of \lazyD{} hypercoercions}

@@

\subsection{All Surely \lazyD{} and Monoidic Cast Representations Respect 
$\mathcal{C}_{\mathtt{D}}$}

@@

We use bisimulation to proof equivalence between the two machines.
\todo[inline]{Which two machines? -JS}
The bisimulation relation of states ($\eqvDS{C}{s}{s}$) is mostly derived from 
the definitions of $ \mathcal{C} $. Exceptions include the relations from 
continuations to pre-continuations, between continuations, and between values. 
We omit subscripts when there is no ambiguity.

@@

\begin{lemma}[Weak Bisimulation between $ \mathcal{S}(\cdot) $ and $ 
	\mathcal{C} $]
	\label{thm:surely-monoidic-reduce}
	For all cast representations $ C $, 
	if $ C $ is surely \lazyD{} and monoidic,
	$ s_1 \in \mathcal{C} $,
	$ s_2 \in \mathcal{S}(C) $, and
	$ s_1 \approx s_2 $, 
	\begin{enumerate}
		\item If $ \judgeCreduce{s_1}{s_3} $,
		then
		there exist $ s_5 \in \mathcal{C} $, $ s_4 \in \mathcal{S}(C) $,
		such that \begin{itemize}
			\item $ s_5 \approx s_4 $, and
			\item $ \judgeSreduce{C}{s_2}{s_4} $,
			\item $ \judgeCreduceTrans{s_3}{s_5} $
		\end{itemize}
		\item If $ \judgeSreduce{C}{s_2}{s_4} $,
		then
		there exist $ s_3 \in \mathcal{C} $,
		such that \begin{itemize}
			\item $ s_3 \approx s_4 $
			\item $ \judgeCreduceTrans{s_1}{s_3} $
		\end{itemize}
	\end{enumerate}
\end{lemma}
\begin{proof}
	See the supplementary material.
\end{proof}

\begin{lemma}[Surely \lazyD\ and Monoidic Cast Representations Respect 
	$ \mathcal{C} $]
	\label{thm:surely-monoidic-eval}
	If $ \judgetype{\emptyset}{e}{T} $ and $ o : T $ and $ C $ is surely 
	\lazyD{} and monoidic,
	\[
	\judgeCeval{e}{o} \; \text{if and only if} \; 
	\judgeSeval{C}{e}{o}
	\]
\end{lemma}
\begin{proof}
  Immediately from Lemma~\ref{thm:surely-monoidic-reduce}.
  \todo[inline]{I'm guessing that you prove this lemma by induction,
    using Lemma~\ref{thm:surely-monoidic-reduce}.
    Proofs by induction are not ``immediate''. -JS}
\end{proof}

\subsection{\lazyD\ Hyper-coercion Respect $ \mathcal{C}_\mathtt{D} $}
\label{sec:LDHC-correct}

>>>>

There are two basic ways to represent cast efficiently: coercions 
\citep{herman2010space}\citep{siek2009exploring}\citep{siek2012interpretations} 
\citep{siek2015blame} and threesomes 
\citep{siek2010threesomes}\citep{garcia2013calculating}. They have different 
strength and 
weakness. \citet{garcia2013calculating} and \citet{siek2015blame} claim that 
coercion is easy to understand, but hard to implement, while threesome is easy 
to implement, but hard to understand. \citet{garcia2013calculating} in addition 
claims that coercion is more difficult to reason about formally, because 
composition of coercions is not structurally recurisive. The developers of 
Grift \citep{kuhlenschmidt2018efficient} affirm that it is tricky to convince 
their proof assistants that composition of coercions terminates.

We present a new cast representation, hypercoercion, which has a structurally 
recursive composition. We believe hypercoercion is easier to understand than 
threesome, because its syntax is close to coercion. For instance, the 
coercion approach represents $ 
\cOOcast{\TOOdyn}{l}{\cOOcast{\POOunit}{l'}{\TOOdyn}} $ as $ \POOunit 
?^{l} ; \POOunit\! $, while the hypercoercion approach gives $ 
\hyperCoercionC{?^l}{\POOunit}{!} $. In Section~\ref{sec:background}, we review 
coercion together with cast calculus, the intermediate languages in 
\citet{kuhlenschmidt2018efficient}. And we present hypercoerion in 
Section~\ref{sec:hypercoercion-definition}.

Another strength of hypercoercion over coercion is that hypercoercion might 
have a more compact memory representations ...
\todo[inline]{Could Jeremy explain here why hypercoercion is more compact in 
memory? -KC}

Many of the existing research on coercions or threesomes include proof(s) of 
correctness, that is, the cast calculus reduces to an observation ($ e 
\Downarrow o $) if and only if their calcului reduce to the same observation ($ 
e \Downarrow_{*} o $), where an observation is a value constructor or a blame. 
Despite the striking similarity of these proves, they are all built from ground 
up. We present a framework for proving correctness of cast reprensentation 
(Section~\ref{sec:framework}), and use it to prove the correctness of 
hypercoercion (Section~\ref{sec:hypercoercion-correctness}). We are optimistic 
that this framework is general enough to include coercion and threesome.

Some readers might know that there are four blame strategies: $ \{Lazy, Eager\} 
\times \{D, UD\} $. Eager strategies detect more potential type errors, but 
blame more programs. $ D $ and $ UD $ assign blame labels differently.
We focus on lazy strategies because recently \citet{new2019gradual} 
show that \eager{} strategies are incompatible with $\eta$-equivalence of 
functions, which suggest that they are not very ideal. 
\todo[inline]{Why do we need to explain why both LD and LUD? -KC}

Here are our major contribution:
\begin{itemize}
	\item hypercoercion, a new space-efficient cast representation, which has a 
	structurally recursive composition and has a more compact memory 
	representation.
	\item a framework in Agda for proving correctness of cast represantations. 
	We conjecture that both coercion and threesome can fit into this framework.
	\item a formal proof that hypercoercion respect the semantics of cast 
	calculus
\end{itemize}

%Around 2006, many work on integrating dynamic typing and static typing emerge, 
%notably gradual types \citep{siek2006gradual} and hybrid types 
%\citep{flanagan2006hybrid}. 
%Later \citet{wadler2009well} introduces Blame Calculus, an intermediate 
%language for gradual types and hybrid types. 
%Blame Calculus throws a blame label when a runtime type-check fails.
%Labels provide useful debugging information, therefore are necessary for 
%practical use of gradually typed languages.
%
%Implementing gradually typed languages naively on top of blame calculus, 
%however, suffices sever space leak.
%For example, translation to Blame Calculus can wrap some expressions with code 
%that performs runtime type-checking (casts), including expressions at 
%tail positions. Thus, mutually tail-recursive functions in source language can 
%consume unbounded space at runtime. There are more examples of space leaking 
%in 
%\citet{herman2010space}.
%
%In 2007, \citet{herman2010space} proposes a solution to this problem by 
%representing casts with coercion \citep{henglein1994dynamic}, 
%which can be composed and normalized. Every coercion looks like a list of 
%trees. For normal coercion, the length of the list is bounded by a small 
%constant. And the depths of trees do not grow during composition and
%normalization. These two facts implies that coercion is a space-efficient cast 
%representation. This coercion, however, does not support blame 
%tracking. 
%
%After that, many efforts have been made to combine blame tracking and 
%space efficiency. 
%
%\citet{siek2009exploring} present another intermediate langauge for gradually 
%typed language, cast calculus \castCalculus\ and achieve both blame-tracking 
%and space-efficiency for this language by decorating \citet{herman2010space}'s 
%coercions with labels. \castCalculus\ is simply typed lambda calculus extended 
%with the dynamic type and cast expressions. The labels in \castCalculus{}, 
%unlike their counterparts in the Blame Calculus, have no polarity. 
%They also propose that there are four blame strategies for \castCalculus{}:
%$ \{Lazy, Eager\} \times \{D, UD\} $. Eager strategies detect more potential 
%type errors, but blame more programs. $ D $ and $ UD $ assign blame labels 
%differently. They claim that their \lazyUD\ cast calculus is the same as the 
%Blame Calculus, modulo label polarity, and prove that their their \lazyUD\ 
%coercion calculus simulates their \lazyUD{} cast calculus. It is unknown, 
%however, whether the simulation in the other direction also works.
%\todo[inline]{Is it important to this paper whether there is a
%simulation in the other direction? -JS
%\\ --- \\
%I guessed to say coercion is correct the other direction is 
%necessary. -KC}
%
%\citet{siek2010threesomes} proposes another approach to combine blame 
%tracking and space efficiency. Their solution represents casts with 
%threesomes, 
%a novel cast representation. Threesome is proved space-efficient as well.
%They prove that their calculus bisimulates the \lazyUD{} coercion 
%calculus in \citet{siek2009exploring} and the Blame Calculus in 
%\citet{wadler2009well}.
%
%\citet{siek2012interpretations} present a revised coercion calculi.
%They simplify the coercion calculi by only 
%working with normal coercions (i.e. coercions in normal form). 
%They conjecture that their revised coercion calculi bisimulate the 
%corresponding cast calculi.
%
%\citet{garcia2013calculating} develops
%threesome for other blame strategies. After studying in depth the connection 
%between coercion and threesome,
%he claims that normal coercion with labels is easy to understand but hard to 
%implement and hard to reason about, and that 
%threesome with labels, however, is easy to implement and easy to reason about 
%but hard to understand. 
%His claim is mostly affirmed by the group of people who develop threesome 
%(\citet{siek2015blame}).
%His threesome calculi are equivalent to the coercion calculi of
%\citet{siek2009exploring} because the former ones are derived from the latter 
%ones.
%
%\citet{siek2015blame} revisit the coercion-based approach. They revise
%the \lazyUD{} coercion in \citet{siek2012interpretations} and prove
%that their revised coercion calculus bisimulates the \lazyUD{} Blame
%Calculus \citet{wadler2009well}.
%
%Last year, \citet{kuhlenschmidt2018efficient} present Grift, a compiler for a 
%gradually typed language of the same name. 
%This implementation is based on the \lazyD{} coercion in 
%\citet{siek2012interpretations}.
%Their result suggests that normal coercion is easy enough to implement in a 
%compiler. 
%
%So far, it seems that normal coercion is the best way to combine 
%space-efficiency and blame-tracking: normal coercion is claimed easy to 
%understand \citep{garcia2013calculating}\citep{siek2015blame} and is shown 
%easy enough to implement in a compiler \citep{kuhlenschmidt2018efficient}.
%Normal coercion is still not satisfactory, however, in at least two aspects.
%Firstly, as we mentioned above, it is hard to reason about normal coercion. 
%The 
%major difficulty is from its non-structurally recursive composition.
%For instance, the developers of Grift report it is tricky to convince
%their proof assistants that the composition terminates.
%
%Secondly, the definition of normal coercion lies unnecessarily on top 
%of coercion: all coercions are normal when they are initially constructed, and 
%\citet{siek2012interpretations} have shown an function that 
%composes and produces normal coercions.
%%
%\todo[inline]{Many researchers would view this as a positive thing,
%  not a negative. For example, the notion of normal forms comes up
%  over and over again in mathematics and logic. -JS
%\\ --- \\
%I guess having a notion of normal forms is more convenient when a 
%notion of normalization is convenient. But we don't even need a notion of 
%normalization for cast representations. -KC
%\\ --- \\
%FWIW, I agree with you. But the point is that the argument is not
%likely to be convincing to the reader, so we're probably better off
%not wasting space on it. -JS
%}
%
%Perhaps unsurprisingly, threesome has a structurally recursive compose and a 
%self-standing definition. Together with \citet{garcia2013calculating}'s 
%work, they suggest that there might be a cast representation whose definition 
%is self-standing, and whose composition is structurally recursive. 
%supercoercion introduced in \citet{garcia2013calculating} could be a promising 
%candidate. It is believed even more complicated than threesome 
%\citep{siek2015blame}. In fact, it uses 10 constructors to deal with an 
%elementary type system with only base types and 
%function types. And four constructors are directly related to function types. 
%Thus, supercoercion might not scale very well to more sophisticated type 
%systems. 
%
%We present hypercoercion, a cast representation whose composition is 
%structurally recursive, and whose definition is self-standing. What's more, it 
%should be at 
%least as easy to implement and understand as normal coercions, because there 
%is 
%a clear connection between them, as we will show in Section 
%\ref{sec:hypercoercion}.
%Our hypercoercion considers sum types and product types, which are not 
%considered in all proofs above. Adding each of them requires us to add only 
%one 
%new constructor to a component of hypercoercion. This suggests that 
%hypercoercion might scale better than supercoercion.
%
%We prove formally that the \lazyD{} (resp. \lazyUD{}) hypercoercion calculus 
%bisimulates the \lazyD{} (resp. \lazyUD{}) cast calculus. This is 
%the first bisimulation proof for the \lazyD{} cast calculus as far as we know.
%We care more about lazy strategies because recently, \citet{new2019gradual} 
%show that \eager{} strategies are incompatible with $\eta$-equivalence of 
%functions, which suggest that these strategies are not very ideal. 
%\todo[inline]{Explain why we care about \lazyD{} and \lazyUD{}. -JS
%  \\ --- \\ Explained, not sure if it suffice. -KC
%  \\ --- \\
%  That's good that you explained why lazy instead of eager.
%  (Also, you should add that we are interested in lazy because
%  that's what Grift does.)\\
%  But you haven't explained why we are interested in both D and UD. -JS
%}
%
%The structure of this paper is as follows. 
%Section \ref{sec:blame-calculus} reviews the state-of-art of $Lazy$ cast 
%calculi. 
%Section \ref{sec:coercion-calculus} reviews the state-of-art of $Lazy$ 
%Coercion 
%Calculi. 
%In section \ref{sec:hypercoercion} we present hypercoercion.
%Section \ref{sec:conclude} concludes.
%\todo[inline]{Lots of missing sections. -JS}

\section{Background} \label{sec:background}

@@

\section{Definition of Hypercoercion} \label{sec:hypercoercion-definition}

@@

\section{a framework for proving correctness of cast representations}
\label{sec:framework}

@@

\section{Correctness Proof of \lazyD{} hypercoercions}
\label{sec:hypercoercion-correctness}

@@

\subsection{All Surely \lazyD{} and Monoidic Cast Representations Respect 
\ineffCEKD}

@@

We use bisimulation to proof equivalence between the two machines.
\todo[inline]{Which two machines? -JS}
The bisimulation relation of states ($\eqvDS{C}{s}{s}$) is mostly derived from 
the definitions of \ineffCEKD. Exceptions include the relations from 
continuations to pre-continuations, between continuations, and between values. 
We omit subscripts when there is no ambiguity.

@@

\begin{lemma}[Weak Bisimulation between $ \mathcal{S}(\cdot) $ and \ineffCEKD]
	\label{thm:surely-monoidic-reduce}
	For all cast representations $ C $, 
	if $ C $ is surely \lazyD{} and monoidic,
	$ s_1 \in \text{\ineffCEKD} $,
	$ s_2 \in \mathcal{S}(C) $, and
	$ s_1 \approx s_2 $, 
	\begin{enumerate}
		\item If $ \judgeDreduce{s_1}{s_3} $,
		then
		there exist $ s_5 \in \text{\ineffCEKD} $, $ s_4 \in \mathcal{S}(C) $,
		such that \begin{itemize}
			\item $ s_5 \approx s_4 $, and
			\item $ \judgeSreduce{C}{s_2}{s_4} $,
			\item $ \judgeDreduceTrans{s_3}{s_5} $
		\end{itemize}
		\item If $ \judgeSreduce{C}{s_2}{s_4} $,
		then
		there exist $ s_3 \in \mathcal{C} $,
		such that \begin{itemize}
			\item $ s_3 \approx s_4 $
			\item $ \judgeDreduceTrans{s_1}{s_3} $
		\end{itemize}
	\end{enumerate}
\end{lemma}
\begin{proof}
	See the supplementary material.
\end{proof}

\begin{lemma}[Surely \lazyD\ and Monoidic Cast Representations Respect 
	$ \mathcal{C} $]
	\label{thm:surely-monoidic-eval}
	If $ \judgetype{\emptyset}{e}{T} $ and $ o : T $ and $ C $ is surely 
	\lazyD{} and monoidic,
	\[
	\judgeDeval{e}{o} \; \text{if and only if} \; 
	\judgeSeval{C}{e}{o}
	\]
\end{lemma}
\begin{proof}
  Immediately from Lemma~\ref{thm:surely-monoidic-reduce}.
  \todo[inline]{I'm guessing that you prove this lemma by induction,
    using Lemma~\ref{thm:surely-monoidic-reduce}.
    Proofs by induction are not ``immediate''. -JS}
\end{proof}

\subsection{\lazyD\ Hyper-coercion Respect \ineffCEKD}
\label{sec:LDHC-correct}
